Deployable Probabilistic Programming

### 2019-04-18 paper/paper.tex
cited onward

	Our attitude differs from the established approach in that
	instead of proposing yet another probabilistic programming
	language, either genuinely different~\cite{MMR+07,Stan17} or
	derived from an existing general-purpose programming
	language by extending the syntax or changing the
	semantics~\cite{GMR+08,TMY+16,GXG18}, we advocate enabling
	probabilistic inference on models written in a
	general-purpose probabilistic programming language, the same
	one as the language used to program the bulk of the system.
	We formulate guidelines for such implementation, and based
	on the guidelines, introduce a probabilistic programming
	facility for the Go programming language~\cite{Golang}. By
	explaining our design choices and implementation techniques,
	we argue that a similar facility can be added to any modern
	general-purpose purpose programming language, giving a
	production system modeling and inference capabilities which
	do not fall short from and sometimes exceed those of
	probabilistic programming-centric systems with custom
	languages and runtimes.

@@

Probabilistic programming is traditionally associated with
design and implementation of probabilistic programming
\textit{languages}. An apparent justification for a new
programming language is that a probabilistic program has a
different semantics~\cite{GHNR14} than a `regular' program of
similar structure. Goodman and Stuhlm\"uller~\cite{GS15} offer a
systematic approach to design and implementation of
probabilistic programming languages based on extension of the
syntax of a (subset of) general-purpose programming language and
transformational compilation to continuation-passing form. Van
de Meent et al~\cite{MPY+18} give a comprehensive overview of
available choices of design and implementation of first-order
and higher-order probabilistic programming languages.
Stan~\cite{Stan17} is built around an imperative probabilistic
programming language with program structure tuned for most
efficient execution of inference on the model.
SlicStan~\cite{GGS19} is built on top of Stan as a
source-to-source compiler, providing the model developer with a
more intuitive language while retaining performance benefits of
Stan. Our work differs from existing approaches in that we
advocate enabling probabilistic programming in a general-purpose
programming language by leveraging existing capabilities of the
language, rather than building a new language on top or besides
an existing one.

@@

Automatic differentiation is widely employed in machine
learning~\cite{BPR+17,MBB+18} where it is also known as
`differentiable programming', and is responsible for enabling
efficient inference in many probabilistic programming
systems~\cite{Stan17,GXG18,ISF+18,BCJ+19,THS+17}. Different
automatic differentiation techniques~\cite{GW08} allow 
different compromises between flexibility, efficiency, and
feature-richness~\cite{SDT14,PGC+17,ISF+18}.  Automatic
differentiation is usually implemented through either
operator overloading~\cite{Stan17,PCG+17,GXG18} or source
code transformation~\cite{ISF+18,MMW18}. Our work, too, relies
on automatic differentiation, implemented as source code
transformation. However, a novelty of our approach is that
instead of using explicit calls or directives to denote parts of
code which needs to be differentiated, we rely on the type
system of Go to selectively differentiate the code relevant for
inference, thus combining advantages of both operator
overloading and source code transformation.

@@

\paragraph{Integration and deployment} Deployment of software
systems is a delicate process involving automatic builds and
maintenance of dependencies. Adding a component, which possibly
introduces additional software dependencies or even a separate
runtime, complicates deployment. Minimizing the burden of
probabilistic programming on the integration and deployment
process should be a major consideration in design or selection
of probabilistic programming tools. Probabilistic programming
systems that are implemented or provide an interface in a
popular programming language, e.g.  Python
(Edward~\cite{THS+17}, Pyro~\cite{BCJ+19}) are easier to
integrate and deploy, however the smaller the footprint of a
probabilistic system, the easier is the adoption.

@@

In Infergo, the inference is performed on a model. An Infergo
model is a data type and methods defined on the data type.
Inference algorithms accept a model object as an argument.
A model object usually encapsulates observations --- the data 
to which the parameter values are adapted.

>>>>

	Our attitude differs from the established approach in that
	instead of proposing yet another probabilistic programming
	language, either genuinely different~\cite{MMR+07,Stan17} or
	derived from an existing general-purpose programming
	language by extending the syntax or changing the
	semantics~\cite{GMR+08,TMY+16,GXG18}, we advocate enabling
	probabilistic inference on models written in a
	general-purpose probabilistic programming language, the same
	one as the language used to program the bulk of the system. 
	We formulate guidelines for such implementation, and based
	on the guidelines, introduce a probabilistic programming
	facility for the Go programming language~\cite{Golang}. By
	explaining our design choices and implementation techniques,
	we argue that a similar facility can be added to any modern
	general-purpose purpose programming language, giving a
	production system modeling and inference capabilities which
	do not fall short from and sometimes exceed those of
	probabilistic programming-centric systems with custom
	languages and runtimes. Availability of such facilities
	in major general-purpose programming languages would free
	probabilistic programming researchers and practitioners
	alike from language wars~\cite{SH14} and foster practical
	applications of probabilistic programming.

@@

Probabilistic programming is traditionally associated with
design and implementation of probabilistic programming
\textit{languages}. An apparent justification for a new
programming language is that a probabilistic program has a
different semantics~\cite{GHNR14} than a `regular' program of
similar structure. \citet{GS15} offer a
systematic approach to design and implementation of
probabilistic programming languages based on extension of the
syntax of a (subset of) general-purpose programming language and
transformational compilation to continuation-passing form. 
\citet{MPY+18} give a comprehensive overview of
available choices of design and implementation of first-order
and higher-order probabilistic programming languages.
Stan~\cite{Stan17} is built around an imperative probabilistic
programming language with program structure tuned for most
efficient execution of inference on the model.
SlicStan~\cite{GGS19} is built on top of Stan as a
source-to-source compiler, providing the model developer with a
more intuitive language while retaining performance benefits of
Stan. Our work differs from existing approaches in that we
advocate enabling probabilistic programming in a general-purpose
programming language by leveraging existing capabilities of the
language, rather than building a new language on top or besides
an existing one.

@@

Automatic differentiation is widely employed in machine
learning~\cite{BPR+17,MBB+18} where it is also known as
`differentiable programming', and is responsible for enabling
efficient inference in many probabilistic programming
systems~\cite{Stan17,GXG18,ISF+18,BCJ+19,THS+17}. Different
automatic differentiation techniques~\cite{GW08} allow 
different compromises between flexibility, efficiency, and
feature-richness~\cite{SDT14,PGC+17,ISF+18}.  Automatic
differentiation is usually implemented through either
operator overloading~\cite{Stan17,PGC+17,GXG18} or source
code transformation~\cite{ISF+18,MMW18}. Our work, too, relies
on automatic differentiation, implemented as source code
transformation. However, a novelty of our approach is that
instead of using explicit calls or directives to denote parts of
code which needs to be differentiated, we rely on the type
system of Go to selectively differentiate the code relevant for
inference, thus combining advantages of both operator
overloading and source code transformation.

@@

\paragraph{Integration and deployment} Deployment of software
systems is a delicate process involving automatic builds and
maintenance of dependencies. Adding a component, which possibly
introduces additional software dependencies or even a separate
runtime, complicates deployment. Minimizing the burden of
probabilistic programming on the integration and deployment
process should be a major consideration in design or selection
of probabilistic programming tools. Probabilistic programming
systems that are implemented or provide an interface in a
popular programming language, e.g.  Python
(Edward~\cite{THS+17}, Pyro~\cite{BCJ+19}) are easier to
integrate and deploy, however the smaller the footprint of a
probabilistic system, the easier is the adoption.
An obstacle in integration of probabilistic programming lies
also in adoption of the probabilistic programming
language~\cite{MR12}, if the latter differs from the
implementation language of the rest of the system.

@@

In Infergo, the inference is performed on a model. Just like in
Stan~\cite{Stan17} models, latent random variables in Infergo
models are continuous. Additionally, Infergo can express and
perform inference on non-deterministic models by including
stochastic choices in the model code.

An Infergo model is a data type and methods defined on the data
type.  Inference algorithms accept a model object as an
argument.  A model object usually encapsulates observations ---
the data to which the parameter values are adapted.
